---
title: "STAT 331 Portfolio"
author: "Kyrene Angelica Tigas"
format: html 
embed-resources: true
layout: margin-left
editor: visual
execute: 
  eval: false
  echo: true
---

[**My Grade:**]{.underline} I believe my grade equivalent to course work evidenced below to be an **A**.

[**Learning Objective Evidence:**]{.underline} In the code chunks below, provide code from Lab or Challenge assignments where you believe you have demonstrated proficiency with the specified learning target. Be sure to specify **where** the code came from (e.g., Lab 4 Question 2).

## Working with Data

**WD-1: I can import data from a *variety* of formats (e.g., csv, xlsx, txt, etc.).**

-   `csv` Example 1

```{r}
#| label: wd-1-csv-1

# Lab 2 Question 1

surveys <- read_csv(here::here("week2", "lab2", "surveys.csv"))
```

-   `csv` Example 2

```{r}
#| label: wd-1-csv-2

# Challenge 3 Question 0

teacher_evals <- read_csv(here("week3","lab3","teacher_evals.csv"))
```

-   `xlsx`

```{r}
#| label: wd-1-xlsx
 
# Practice Activity 4 - Military Spending

military <- read_xlsx("gov_spending_per_capita.xlsx",
                      sheet = "Share of Govt. spending",
                      skip  = 7,
                      n_max = 190,
                      na = c(". .", "xxx", "..")
                      )
```

**WD-2: I can select necessary columns from a dataset.**

-   Example selecting specified columns

```{r}
#| label: wd-2-ex-1

# Challenge 3 Question 1

teacher_evals_compare <- teacher_evals |> 
  filter(question_no == '903') |>
  mutate(set_level = if_else(SET_score_avg >= 4, "excellent", "standard" )) |> 
  mutate(sen_level = case_when(seniority <= 4 ~ "junior",
                                seniority <= 8 ~ "senior",
                                seniority > 8 ~ "very senior")) |> 
  select(course_id, set_level, sen_level)


# What I did:
# I filtered a teacher evaluation dataset for specific questions and created new categorical variables. 
# After these transformations, I used select() to keep only the three most relevant columns for analysis. This shows targeted column selection after complex data manipulation
```

-   Example removing specified columns

```{r}
#| label: wd-2-ex-2

# Lab 4 Question 2 

# Success comment: Nice work filtering first!

# ORIGINAL
ca_childcare <- counties |>
  filter(state_abbreviation == "CA") |>
  inner_join(childcare_costs, by = join_by(county_fips_code))

# MODIFIED 
ca_childcare <- counties |> 
  filter(state_abbreviation == "CA") |> 
  select(-state_name) |>  
  inner_join(childcare_costs, by = "county_fips_code")

# What I did: 
# I removed redundant columns when filtering for California counties to create cleaner datasets. 
# Using select(-state_name) eliminated unnecessary information since all rows were from California. 
# This demonstrates strategic column removal to streamline data.
```

-   Example selecting columns based on logical values (e.g., `starts_with()`, `ends_with()`, `contains()`, `where()`)

```{r}
#| label: wd-2-ex-3

# Challenge 4 Filter Bay Area

# ORIGINAL
bay_area_long <- bay_area_childcare |>
  pivot_longer(
    cols = c(mc_infant, mc_toddler, mc_preschool, 
             mfcc_infant, mfcc_toddler, mfcc_preschool),
    names_to = "type_age",
    values_to = "median_price"
  )


# Midterm comment: include starts_with()

# REVISION
bay_area_long <- bay_area_childcare |>
  pivot_longer(
    cols = starts_with("mc_") | starts_with("mfcc_"),
    names_to = "type_age",
    values_to = "median_price"
  )


# What I did:
# I improved maintainability by using starts_with() for dynamic column selection instead of listing each column individually. 
# This approach adapts automatically if the dataset structure changes. The revision shows more robust and flexible column selection techniques
```

**WD-3: I can filter rows from a dataframe for a *variety* of data types (e.g., numeric, integer, character, factor, date).**

-   Numeric Example 1

```{r}
#| label: wd-3-numeric-ex-1

# Lab 4 Question 5 

# Growing: Notice that your table still has a grouping variable! Because you grouped by **two** variables, your first group was not dropped! Can you give your column names better titles, so the reader knows what "2008" and "2018" represent? What are the values stored in these columns?

# ORIGINAL
ca_childcare |>
  filter(study_year %in% c(2008, 2018)) |>
  group_by(region, study_year) |>
  summarise(median_income = median(mhi_2018, na.rm = TRUE)) |>
  pivot_wider(names_from = study_year, values_from = median_income) |>
  arrange(desc(`2018`))


# REVISED
income_comparison <- ca_childcare |>
  filter(study_year %in% c(2008, 2018)) |>
  group_by(region, study_year) |>
  summarise(median_income = median(mhi_2018, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(
    names_from = study_year, 
    values_from = median_income,
    names_prefix = "median_income_"
  ) |>
  arrange(desc(median_income_2018))


# What I did: 
# I fixed two issues in the code. First, I added .groups = "drop" to remove the grouping structure after summarizing. Second, I used names_prefix to create clearer column names that show these are median income values for each year, making the table easier to understand.
```

-   Numeric Example 2

```{r}
#| label: wd-3-numeric-ex-2

# Challenge 3 Question 1 - Filtering question number

teacher_evals_compare <- teacher_evals |> 
  filter(question_no == '903') |>
  mutate(set_level = if_else(SET_score_avg >= 4, "excellent", "standard" )) |> 
  mutate(sen_level = case_when(seniority <= 4 ~ "junior",
                                seniority <= 8 ~ "senior",
                                seniority > 8 ~ "very senior")) |> 
  select(course_id, set_level, sen_level)


# What I did:
# I isolated survey responses for question 903 by filtering based on the question number variable. 
# This targeted filtering created a subset focused on a specific aspect of the teacher evaluations. 
# The approach shows precise numeric filtering for analytical purposes.
```

-   Character Example 1 (any context)

```{r}
#| label: wd-3-character

# Challenge 4 Filter Bay Area

bay_area_counties <- c("Alameda","Contra Costa","Marin","San Francisco","San Mateo","Santa Clara","Solano","Napa")

bay_area_childcare <- ca_childcare %>%
  filter(county_name %in% bay_area_counties)

# Pivot data to long format for plotting
bay_area_long <- bay_area_childcare %>%
  pivot_longer(cols = c(mc_infant, mc_toddler, mc_preschool,
                        mfcc_infant, mfcc_toddler, mfcc_preschool),
               names_to = "type_age",
               values_to = "median_price") %>%
  mutate(
    age_group = case_when(
      str_detect(type_age, "infant") ~ "Infant",
      str_detect(type_age, "toddler") ~ "Toddler",
      str_detect(type_age, "preschool") ~ "Preschool"
    ),
    childcare_type = case_when(
      str_detect(type_age, "^mc_") ~ "Center-Based",
      str_detect(type_age, "^mfcc_") ~ "Family-Based"
    ),
    pct_families_of_color = 1 - (one_race_w / households),
    pct_white_families = one_race_w / households
  )

# What I did:
# I filtered California counties to include only those in the Bay Area region. 
# Using the %in% operator with a predefined character vector, I created a focused geographic subset. 
# This demonstrates effective character-based filtering for regional analysis.
```

-   Character Example 2 (example must use functions from **stringr**)

```{r}
#| label: wd-3-string

# Challenge 4 Filter Bay Area

bay_area_long <- bay_area_childcare |>
  pivot_longer(cols = c(mc_infant, mc_toddler, mc_preschool,
                        mfcc_infant, mfcc_toddler, mfcc_preschool),
               names_to = "type_age",
               values_to = "median_price") |>
  mutate(
    age_group = case_when(
      str_detect(type_age, "infant") ~ "Infant",
      str_detect(type_age, "toddler") ~ "Toddler",
      str_detect(type_age, "preschool") ~ "Preschool"
    ),
    childcare_type = case_when(
      str_detect(type_age, "^mc_") ~ "Center-Based",
      str_detect(type_age, "^mfcc_") ~ "Family-Based"
    ),
    pct_families_of_color = 1 - (one_race_w / households),
    pct_white_families = one_race_w / households
  )


# What I did:
# I used str_detect() to identify patterns in column names and create meaningful categorical variables. 
# This string-based filtering enabled me to extract age group and childcare type information from complex column names. 
# The approach shows sophisticated string manipulation for data categorization.
```

-   Date (example must use functions from **lubridate**)

```{r}
#| label: wd-3-date

# Lab 7 Question 3

# The fish dataset analysis originally focused on year as numeric, not dates

# ORIGINAL

fish |>
  filter(year >= 1990) |>  # Simple numeric filtering only
  mutate(
    season = if_else(year %% 2 == 0, "even_year", "odd_year")  
  ) |>
  select(year, species, length, weight)


# MODIFIED to use lubridate

library(lubridate)

fish_with_dates <- fish_data |>
  mutate(
    capture_date = ymd(capture_date),
    is_summer = month(capture_date) %in% 6:8   

fish_summer_captures <- fish_with_dates |>
  filter(
    month(capture_date) >= 6,
    month(capture_date) <= 8,
    year(capture_date) >= 1990
  ) |>
  select(year, species, length, weight, capture_date, is_summer)


# What I did:
# I transformed simple numeric years into full date objects as the original code only filtered by numeric year values. The revision creates actual Date objects with make_date(), extracts components with month() and year(), and filters using these date functions. 
```

**WD-4: I can modify existing variables and create new variables in a dataframe for a *variety* of data types (e.g., numeric, integer, character, factor, date).**

-   Numeric Example 1

```{r}
#| label: wd-4-numeric-ex-1

# Challenge 3 Question 1 

# ORIGINAL
teacher_evals_compare <- teacher_evals |> 
  filter(question_no == '903') |>
  mutate(set_level = if_else(SET_score_avg >= 4, "excellent", "standard" )) |> 
  mutate(sen_level = case_when(seniority <= 4 ~ "junior",
                                seniority <= 8 ~ "senior",
                                seniority > 8 ~ "very senior")) |> 
  select(course_id, set_level, sen_level)

# Comment in midterm rubric: multiple calls to mutate

# REVISED 
teacher_evals_compare <- teacher_evals |> 
  filter(question_no == '903') |>
  mutate(
    set_level = if_else(SET_score_avg >= 4, "excellent", "standard"),
    sen_level = case_when(
      seniority <= 4 ~ "junior",
      seniority <= 8 ~ "senior",
      seniority > 8 ~ "very senior"
    )
  ) |> 
  select(course_id, set_level, sen_level)

# What I did: 
# I combined two separate mutate() calls into one. The original
# had mutate() twice which is inefficient. This single mutate() does both 
# transformations at once, making the code cleaner and more efficient.
```

-   Numeric Example 2

```{r}
#| label: wd-4-numeric-ex-2

# Challenge 7 Question 1

fish <- fish |>
  mutate(length_cm = length/10)

# What I did: 
# I created a new length variable by dividing existing measurements by 10 to convert millimeters to centimeters. 
# This simple arithmetic transformation demonstrates basic variable creation for unit conversion. The approach shows practical data transformation skills. 
```

-   Factor Example 1 (renaming levels)

```{r}
#| label: wd-4-factor-ex-1

# Lab 4 Question 4

# Growing comment: You should only call the mutate() function once!

# ORIGINAL

ca_childcare <- ca_childcare |> 
  mutate(county_name = str_remove(county_name, " County")) |> 
  mutate(region = fct_collapse(county_name,
    "Superior California" = c("Butte", "Colusa", "El Dorado", "Glenn", "Lassen", "Modoc", "Nevada", "Placer", "Plumas", "Sacramento", "Shasta", "Sierra", "Siskiyou", "Sutter", "Tehama", "Yolo", "Yuba"),
    "North Coast" = c("Del Norte", "Humboldt", "Lake", "Mendocino", "Napa", "Sonoma", "Trinity"),
    "San Francisco Bay Area" = c("Alameda", "Contra Costa", "Marin", "San Francisco", "San Mateo", "Santa Clara", "Solano"),
    "Northern San Joaquin Valley" = c("Alpine", "Amador", "Calaveras", "Madera", "Mariposa", "Merced", "Mono", "San Joaquin", "Stanislaus", "Tuolumne")
  ))
    

# REVISED

ca_childcare <- ca_childcare |> 
  mutate(
    county_name = str_remove(county_name, " County"),
    region = fct_collapse(county_name,
      "Superior California" = c("Butte", "Colusa", "El Dorado", "Glenn", "Lassen", "Modoc", "Nevada", "Placer", "Plumas", "Sacramento", "Shasta", "Sierra", "Siskiyou", "Sutter", "Tehama", "Yolo", "Yuba"),
      "North Coast" = c("Del Norte", "Humboldt", "Lake", "Mendocino", "Napa", "Sonoma", "Trinity"),
      "San Francisco Bay Area" = c("Alameda", "Contra Costa", "Marin", "San Francisco", "San Mateo", "Santa Clara", "Solano"),
      "Northern San Joaquin Valley" = c("Alpine", "Amador", "Calaveras", "Madera", "Mariposa", "Merced", "Mono", "San Joaquin", "Stanislaus", "Tuolumne"),
      "Central Coast" = c("Monterey", "San Benito", "San Luis Obispo", "Santa Barbara", "Santa Cruz", "Ventura")
    )
  )

# What I did: 
# I collapsed multiple county names into regional factor categories using fct_collapse().
# I corrected the original code by combining two separate mutate operations into one. 
# This demonstrates efficient factor level manipulation with geographic data.
```

-   Factor Example 2 (reordering levels)

```{r}
#| label: wd-4-factor-ex-2

# Lab 4 Question 7 

# Growing comment: Nice work pivoting and modifying the age variable.
# Growing comment: Nice work reordering your legend and your facets!!

ca_childcare_long <- ca_childcare |> 
  pivot_longer(cols = mc_infant:mc_preschool,
                names_to = "age_group",
                values_to = "median_price") |> 
  select(region, study_year, median_price, age_group) |> 
  mutate(age_group = case_when(age_group == "mc_infant" ~ "Infant",
                                age_group == "mc_toddler" ~ "Toddler",
                                age_group == "mc_preschool" ~ "Preschool"),
          age_group = as.factor(age_group),
          age_group = factor(age_group, levels = c("Infant", "Toddler", "Preschool")))


# What I did: 
# I reordered factor levels to control their display sequence in visualizations and analyses. 
# Using factor() with specified levels ensured that age groups appeared in logical order.
# This shows understanding of factor ordering for presentation purposes.
```

-   Character (example must use functions from **stringr**)

```{r}
#| label: wd-4-string

# Lab 4 Question 4

# ORIGINAL 
ca_childcare <- ca_childcare |> 
  mutate(county_name = str_remove(county_name, " County")) |>
  mutate(region = fct_collapse(county_name,
      "Superior California" = c("Butte", "Colusa", "El Dorado", "Glenn", "Lassen", "Modoc", "Nevada", "Placer", "Plumas", "Sacramento", "Shasta", "Sierra", "Siskiyou", "Sutter", "Tehama", "Yolo", "Yuba"),
      "North Coast" = c("Del Norte", "Humboldt", "Lake", "Mendocino", "Napa", "Sonoma", "Trinity"),
      "San Francisco Bay Area" = c("Alameda", "Contra Costa", "Marin", "San Francisco", "San Mateo", "Santa Clara", "Solano"),
      "Northern San Joaquin Valley" = c("Alpine", "Amador", "Calaveras", "Madera", "Mariposa", "Merced", "Mono", "San Joaquin", "Stanislaus", "Tuolumne"),
      "Central Coast" = c("Monterey", "San Benito", "San Luis Obispo", "Santa Barbara", "Santa Cruz",     "Ventura"),
      "Southern San Joaquin Valley" = c("Fresno", "Inyo", "Kern", "Kings", "Tulare"),
      "Inland Empire" = c("Riverside", "San Bernardino"),
      "Los Angeles County" = c("Los Angeles"),
      "Orange County" = c("Orange"),
      "San Diego - Imperial" = c("Imperial", "San Diego")))
  ca_childcare
  
  
# Growing comment: you should only call the mutate() function once!
  
# REVSION
ca_childcare <- ca_childcare |> 
  mutate(
    county_name = str_remove(county_name, " County"),
    region = fct_collapse(
      county_name,
      "Superior California" = c("Butte", "Colusa", "El Dorado", "Glenn", "Lassen", 
                                "Modoc", "Nevada", "Placer", "Plumas", "Sacramento", 
                                "Shasta", "Sierra", "Siskiyou", "Sutter", "Tehama", 
                                "Yolo", "Yuba"),
      "North Coast" = c("Del Norte", "Humboldt", "Lake", "Mendocino", "Napa", 
                        "Sonoma", "Trinity"),
      "San Francisco Bay Area" = c("Alameda", "Contra Costa", "Marin", 
                                   "San Francisco", "San Mateo", "Santa Clara", 
                                   "Solano"),
      "Northern San Joaquin Valley" = c("Alpine", "Amador", "Calaveras", "Madera", 
                                        "Mariposa", "Merced", "Mono", "San Joaquin", 
                                        "Stanislaus", "Tuolumne"),
      "Central Coast" = c("Monterey", "San Benito", "San Luis Obispo", 
                          "Santa Barbara", "Santa Cruz", "Ventura"),
      "Southern San Joaquin Valley" = c("Fresno", "Inyo", "Kern", "Kings", "Tulare"),
      "Inland Empire" = c("Riverside", "San Bernardino"),
      "Los Angeles County" = c("Los Angeles"),
      "Orange County" = c("Orange"),
      "San Diego - Imperial" = c("Imperial", "San Diego")
    )
  )


# What I did: 
# I cleaned county names by removing the "County" suffix using str_remove(). 
# This string manipulation created cleaner, more standardized geographic identifiers. 
```

-   Date (example must use functions from **lubridate**)

```{r}
#| label: wd-4-date

# Lab 4 Question 3

# ORIGINAL 

ca_childcare <- counties |> 
  filter(state_abbreviation == "CA") |> 
  inner_join(childcare_costs, by = join_by(county_fips_code))

ca_childcare

# MODIFIED to include lubridate functions

# study_year from childcare_costs <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-05-09/childcare_costs.csv')

# lubridate functions from https://r4ds.hadley.nz/datetimes.html#date-time-components

ca_childcare_dates <- ca_childcare |>
  mutate(
    # Create date objects from study_year column 
    study_date = make_date(year = study_year, month = 7, day = 1),  
    decade = floor_date(study_date, years(10)),  
    fiscal_year_start = make_date(year = study_year - 1, month = 7, day = 1),
    fiscal_year_end = make_date(year = study_year, month = 6, day = 30),
    fiscal_year_length = interval(fiscal_year_start, fiscal_year_end) / days(1)
  ) |>
  select(county_name, study_year, study_date, decade, fiscal_year_length) |>
  head(10)

ca_childcare_dates


# What I did:
# I used lubridate functions to transform the numeric study_year into date variables, including study_date, decade groupings, and fiscal year start/end dates. 
# I also calculated the length of each fiscal year in days. 
# This shows how to modify existing variables and create new ones of the date type, demonstrating sophisticated date manipulation
```

**WD-5: I can use mutating joins to combine multiple dataframes.**

-   `left_join()` Example 1

```{r}
#| label: wd-5-left-ex-1

# Left join keeps all rows from the left table and matches from the right

# Lab 4 Question 3

# Success comment: Nice work noticing that there are two keys that need to be used for joining!

ca_childcare <- ca_childcare |>
  left_join(tax_rev, by = join_by(county_name == entity_name, study_year == year))


# What I did:
# I combined childcare data with tax revenue information using a left join based on county name and year.
# This preserved all childcare records while adding available tax data. 
# The approach shows proper join key specification for complex merges.
```

-   `right_join()` Example 1

```{r}
#| label: wd-5-right

# right join keeps all rows from the right and matches from the left

# Lab 4 Question 3 

# Success comment: Nice work noticing that there are two keys that need to be used for joining!

# ORIGINAL 
ca_childcare <- ca_childcare |>
  left_join(tax_rev, by = join_by(county_name == entity_name, study_year == year))

# MODIFIED to use right_join()

tax_rev |>
  right_join(ca_childcare, by = c("entity_name" = "county_name", "year" = "study_year"))

# What I did: 
# I used a right join starting with tax revenue data and joining childcare costs. This keeps all tax records while matching childcare information.
```

-   `left_join()` **or** `right_join()` Example 2

```{r}
#| label: wd-5-left-right-ex-2

# Lab 7 Question 3

# ORIGINAL

fish_missing <- fish |>
  mutate(weight_missing = is.na(weight))

missing_counts <- fish_missing |>
  count(year, section, trip, weight_missing)


# MODIFIED to use left join

area_summary <- missing_counts |>
  group_by(year, section) |>
  summarise(
    total_observations = sum(n),
    missing_observations = sum(n[weight_missing == TRUE]),
    missing_percentage = missing_observations / total_observations * 100,
    .groups = "drop"
  )

enriched_fish <- fish |>
  left_join(area_summary, by = c("year", "section"))


# What I did:
# I calculated summary statistics for missing weights by year and section, 
# then used a left join to add these metrics back to the original fish dataset. 
# This enriched the dataset with area-level data quality information, 
# demonstrating how to combine multiple dataframes using mutating joins.
```

-   `inner_join()` Example 1

```{r}
#| label: wd-5-inner-ex-1 


# inner join keeps only rows that match in both tables

# Challenge 4 - Bay Area Analysis

# ORIGINAL 

bay_area_counties <- c("Alameda","Contra Costa","Marin","San Francisco",
                       "San Mateo","Santa Clara","Solano","Napa")
bay_area_childcare <- ca_childcare |>
  filter(county_name %in% bay_area_counties)


# MODIFIED to use inner_join() instead: 
bay_area_lookup <- tibble(
  county_name = c("Alameda", "Contra Costa", "Marin", "San Francisco",
                  "San Mateo", "Santa Clara", "Solano", "Napa"),
  region_type = "Bay Area"
)

bay_area_childcare_joined <- ca_childcare |>
  inner_join(bay_area_lookup, by = "county_name")

# What I did: 
# I used inner_join() to filter the childcare data to only Bay Area counties while also
# adding region_type information from the lookup table. This demonstrates how inner_join()
# can serve as both a filter and a way to add new columns from another dataset.
```

-   `inner_join()` Example 2

```{r}
#| label: wd-5-inner-ex-2

# inner join keeps only rows that match in both tables

# Lab 4 Question 2

# Success Comment: Nice work filtering first! Technically, you don't need to use join_by() if both datasets have the same column names. You can simply use by = "county_fips_code".

ca_childcare <- counties |> 
  filter(state_abbreviation == "CA") |> 
  inner_join(childcare_costs, by = join_by(county_fips_code))


# REVISED using comment
# I simplified the join syntax by removing the unnecessary join_by() wrapper. Since both datasets shared the same column name, I used the simpler by = "county_fips_code" format instead.

ca_childcare <- counties |> 
  filter(state_abbreviation == "CA") |> 
  inner_join(childcare_costs, by = "county_fips_code")
```

**WD-6: I can use filtering joins to filter rows from a dataframe.**

-   `semi_join()`

```{r}
#| label: wd-6-semi

# semi join keeps rows from the left that have a match in the right but doesn’t add columns from the right.

# Lab 7 Question 3

fish_missing <- fish |>
  mutate(weight_missing = is.na(weight))

missing_counts <- fish_missing |>
  count(year, section, trip, weight_missing)
        
# Using semi_join() to filter problem areas
# MODIFIED

# find year-section combinations with any missing weights
problem_areas <- missing_counts |>
  filter(weight_missing == TRUE) |>
  distinct(year, section)

# use semi_join() to get all fish from problem areas
problem_area_fish <- fish |>
  semi_join(problem_areas, by = c("year", "section"))


# What I did:
# semi_join() is used to filter the fish dataset to only include rows that match specific year and section combinations with missing weight values. 
# Instead of adding new columns, the join acts purely as a filtering tool, 
# allowing me to isolate only the fish records that come from problem areas with missing data.
```

-   `anti_join()`

```{r}
#| label: wd-6-anti

# Lab 4 Question 3

# Success comment: Nice work noticing that there are two keys that need to be used for joining!

ca_childcare <- ca_childcare |>
  left_join(tax_rev, by = c("county_name" = "entity_name",
                            "study_year" = "year"))


## MODIFIED - Finding missing counties:
counties_without_tax <- ca_childcare |>
  filter(study_year == 2018) |>
  distinct(county_name, study_year) |>
  anti_join(tax_rev, 
            by = c("county_name" = "entity_name", "study_year" = "year"))

# Check which counties are missing
counties_without_tax |> 
  distinct(county_name)


# What I did: 
# anti_join() is used to filter out all county–year combinations that have matching tax revenue data, leaving only the counties that are missing tax information for 2018. Rather than merging datasets, the join is used to identify gaps in the data by keeping only unmatched rows.
```

**WD-7: I can pivot dataframes from long to wide and visa versa**

-   `pivot_longer()`

```{r}
#| label: wd-7-long

# Challenge 4 Description of Data

bay_area_long <- bay_area_childcare |>
  pivot_longer(cols = c(mc_infant, mc_toddler, mc_preschool,
                        mfcc_infant, mfcc_toddler, mfcc_preschool),
               names_to = "type_age",
               values_to = "median_price") |>
  mutate(
    age_group = case_when(
      str_detect(type_age, "infant") ~ "Infant",
      str_detect(type_age, "toddler") ~ "Toddler",
      str_detect(type_age, "preschool") ~ "Preschool"
    ),
    childcare_type = case_when(
      str_detect(type_age, "^mc_") ~ "Center-Based",
      str_detect(type_age, "^mfcc_") ~ "Family-Based"
    ),
    pct_families_of_color = 1 - (one_race_w / households),
    pct_white_families = one_race_w / households
  )

# Pivot proportions for plotting families of color vs white
bay_area_plot_data <- bay_area_long |>
  pivot_longer(cols = c(pct_families_of_color, pct_white_families),
               names_to = "household_group",
               values_to = "proportion") |>
  mutate(
    household_group = recode(household_group,
                             "pct_families_of_color" = "Families of Color",
                             "pct_white_families" = "White Families")
  )

# What I did:
# I transformed multiple childcare cost columns into a long format suitable for analysis and visualization. 
# This restructuring converted separate age-specific columns into categorical variables with corresponding values.
```

-   `pivot_wider()`

```{r}
#| label: wd-7-wide

# Lab 4 Q5

# Growing Comment: Notice that your table still has a grouping variable! Because you grouped by **two** variables, your first group was not dropped!
# Growing Comment: Can you give your column names better titles, so the reader knows what "2008" and "2018" represent? What are the values stored in these columns?

ca_childcare |>
  filter(study_year %in% c(2008, 2018)) |>
  group_by(region, study_year) |>
  summarise(median_income = median(mhi_2018, na.rm = TRUE)) |>
  pivot_wider(names_from = study_year, values_from = median_income) |>
  arrange(desc(`2018`))


# REVISION

# I fixed two issues in the code. First, I added .groups = "drop" to remove the grouping structure after summarizing. 
# Second, I used names_prefix to create clearer column names that show these are median income values for each year, making the table easier to understand.
# I transformed summarized median income data from long to wide format for year comparison. Adding names_prefix created clearer column names indicating what the values represent. 

income_comparison <- ca_childcare |>
  filter(study_year %in% c(2008, 2018)) |>
  group_by(region, study_year) |>
  summarise(median_income = median(mhi_2018, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(
    names_from = study_year, 
    values_from = median_income,
    names_prefix = "median_income_"
  ) |>
  arrange(desc(median_income_2018))



```

## Reproducibility

**R-1: I can create professional looking, reproducible analyses using RStudio projects, Quarto documents, and the here package.**

The following assignments satisfy the above criteria:

-   **Lab 2**: Exploring Rodents with ggplot2 - Used `here::here()` package for file paths and created reproducible analysis with embedded resources

-   **Lab 4**: Childcare Costs in California - Demonstrated reproducible data pipelines with proper project organization

-   **Lab 7**: Functions + Fish - Used `here()` package

-   **Lab 8**: Data Frame Functions - Professional Quarto document with proper code-fold, embed-resources, and organized function documentation

-   **Lab 9**: Used `here::here()` for multiple dataset imports and demonstrated professional table formatting with reproducible code

**R-2: I can write well documented and tidy code.**

-   Example of **ggplot2** plotting

```{r}
#| label: r-2-1

# Lab 2 Question 8

# create boxplot of weight by species and sex
ggplot(data = surveys,
        mapping = aes (x = species, y = weight, fill = sex)) +
  
      # add boxplots (hide outliers)
      geom_boxplot(outlier.shape = NA) +
  
      # add individual data points 
      geom_jitter(color = "steelblue", alpha = 0.1) +
  
      # add title and axis labels
      labs(
        title = "Side-by-Side Boxplots with Observations of Body Weight by Species",
        x = "Species",
        y = "Body Weight (grams)"
) +
      
      # adjust plot appearance
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(family = "Times New Roman"),
        plot.title = element_text(size = 16, face = "bold"),
        panel.grid.minor = element_blank(),
        legend.position = "none"
        ) 

# What I did:
# I created a complex boxplot visualization with overlaid jittered points and custom theming. 
# The code includes clear sectioning with comments for different plot components.
```

-   Example of **dplyr** pipeline

```{r}
#| label: r-2-2

# Challenge 3 Question 1 

teacher_evals_compare <- teacher_evals |> 
  
  # keep only responses for question 903 
  filter(question_no == '903') |>
  
  # create a new column labeling SET scores as excellent
  mutate(set_level = if_else(SET_score_avg >= 4, "excellent", "standard" )) |> 
  
  # create a new column grouping instructors by seniority
  mutate(sen_level = case_when(seniority <= 4 ~ "junior",
                                seniority <= 8 ~ "senior",
                                seniority > 8 ~ "very senior")) |> 
  
  # keep only relevant columns for comparison
  select(course_id, set_level, sen_level)



# What I did:
# I constructed a readable data transformation pipeline with logical commenting at each step. 
# The code flows naturally from filtering to mutating to selecting operations. 
```

-   Example of function formatting

```{r}
#| label: r-2-3


# Lab 7 Question 4 

rescale_01 <- function(x) {
  
  # if() from Base R cheatsheet https://rstudio.github.io/cheatsheets/base-r.pdf
  
  # check that the input is numeric 
  if (!is.numeric(x)) {
    return("Input must be numeric")
  }
  
  # check that the vector has more than one value
  if (length(x) <= 1) {
    return("Input vector length must be greater than 1")
  }
  
  # rng() from textbook (25.2)
  # find min and max values (ignore NA)
  rng <- range(x, na.rm = TRUE)
  min_val <- rng[1]
  max_val <- rng[2]
  
  # rescale values to be between 0 and 1 
  (x - min_val) / (max_val - min_val)
}


# What I did :
#  I wrote a robust rescaling function with input validation and clear internal documentation. 
# The function handles edge cases gracefully with informative error messages. 
```

**R-3: I can write robust programs that are resistant to changes in inputs.**

-   Example (any context)

```{r}
#| label: r-3-example

# Challenge 7 Question 2 - Function with input validation

replace_unlikely <- function(measurements, min_reasonable, max_reasonable) {
  if_else(measurements < min_reasonable | measurements > max_reasonable, 
          NA, measurements)
}


# What I did:
# I created a function that replaces unrealistic measurements with NA values based on reasonable ranges. 
# This data cleaning function handles boundary cases systematically. 
```

-   Example (function stops)

```{r}
#| label: r-3-function-stops

# Lab 7 Question 5 

rescale_01 <- function(x) {
  if (!is.numeric(x)) {
    stop("Input must be numeric")
  }
  if (length(x) <= 1) {
    stop("Input vector length must be greater than 1")
  }
  
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}


# What I did:
# I enhanced a rescaling function to stop execution with clear error messages for invalid inputs. 
# Using stop() instead of returning strings makes the function behave more predictably. 
```

## Data Visualization & Summarization

**DVS-1: I can create visualizations for a *variety* of variable types (e.g., numeric, character, factor, date) - MUST BE UNIQUE EXAMPLES**

-   At least two numeric variables

```{r}
#| label: dvs-1-num

# Lab 2 Question 4

ggplot(data = surveys, 
        mapping = aes(x = weight, y = hindfoot_length)) + 
    geom_point(alpha = 0.1) +
    facet_wrap(~species) +
    labs(
      title = "Relationship Between Body Weight and Hindfoot Length by Species", 
      subtitle = "Hindfoot Length (mm)",
      x = "Body Weight (grams)",
      y = ""
    )

# What I did:
# I created a scatterplot examining the relationship between weight and hindfoot length across species. 
# Faceting by species enabled comparison of this relationship across different groups. 
```

-   At least one numeric variable and one categorical variable

```{r}
#| label: dvs-2-num-cat

# Lab 10 Question 3 

# Success comment: Nice work adding the total babies and the number of simulations to the plot title!

enframe(results, value = "correct_babies") |>
  count(correct_babies) |>
  mutate(proportion = n / sum(n)) |>
  ggplot(aes(x = factor(correct_babies), y = proportion)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    x = "Number of Babies Correctly Returned",
    y = "Proportion",
    title = "Distribution of Correct Baby Returns",
    subtitle = "Based on 10,000 simulations with 4 babies"
  ) +
  theme_minimal()

# What I did: 
# I made a random babies simulation bar plot
# The proportion (n for count) is the numeric variable
# correct_babies is the categorical variable
```

-   At least two categorical variables

```{r}
#| label: dvs-2-cat

# Challenge 3 Question 2 

# Success comment:  Incredible work! Nice job matching the colors, the percentage labels, and the legend position! Technically, the percent_format() function has been superseded. Meaning, there are other functions you should use instead. Specifically, the label_percent() function has replaced this function.

ggplot(data = teacher_evals_compare,
       mapping = aes(x = sen_level,
                     fill = set_level)) +
  geom_bar(position = "fill") +
  labs(title = "Evaluation of Teachers' Use of Activities",
       x = "Years of Experience",
       y = "",
       fill = "Evaluation Rating") +
  scale_fill_manual(values = c("standard" = "#D2B48C", "excellent" = '#B19CD9')) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "top",
    panel.grid.major = element_blank(),
  )


# REVISED: Updated percent_format() to label_percent() as recommended


ggplot(data = teacher_evals_compare,
       mapping = aes(x = sen_level,
                     fill = set_level)) +
  geom_bar(position = "fill") +
  labs(title = "Evaluation of Teachers' Use of Activities",
       x = "Years of Experience",
       y = "",
       fill = "Evaluation Rating") +
  scale_fill_manual(values = c("standard" = "#D2B48C", "excellent" = '#B19CD9')) +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "top",
    panel.grid.major = element_blank(),
  )


# What I did:
# I created a stacked bar plot showing evaluation ratings across different experience levels. 
# Using position="fill" converted counts to proportions for fair comparison across groups.
```

-   Dates (time series plot)

```{r}
#| label: dvs-2-date

# Lab 4 Question 7


# Growing comment: Do you need to rotate the x-axis labels? Is there a way you can make this plot without making people tilt their head to read the titles of your plot?

# Growing: Your text is quite squished! The size of the text in my plot is not the default size, because it ends up being squished. Can you figure out how to resize the size of the text used in your plot?

ggplot(data = ca_childcare_long,
       aes(x = study_year, y = median_price, 
           color = fct_reorder2(region, study_year, median_price))) +
  geom_point() +
  facet_grid(cols = vars(age_group)) +
  geom_smooth(method = "loess") +
  scale_x_continuous(breaks = seq(2008, 2018, 2)) +
  labs(x = "Study Year",
       y = " ",
       subtitle = "Weekly Median Price for Center-Based Childcare ($)",
       color = "California Region")



# REVISION

ggplot(data = ca_childcare_long,
       aes(x = study_year, y = median_price, 
           color = fct_reorder2(region, study_year, median_price))) +
  geom_point() +
  geom_smooth(method = "loess") +
  facet_grid(cols = vars(age_group)) +
  scale_x_continuous(breaks = seq(2008, 2018, 2)) +
  labs(x = "Study Year",
       y = "Weekly Median Price ($)",
       subtitle = "Weekly Median Price for Center-Based Childcare ($)",
       color = "California Region") +
  theme(
    axis.text.x = element_text(hjust = 0.5, size = 12),  # horizontal labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

# What I did:
# I Created a time series visualization of weekly median childcare prices in California from 2008 to 2018, using study_year (date/numeric) on the x-axis and median_price (numeric) on the y-axis. The plot included facets for different age groups (factor variable) and colored lines for each region, demonstrating the ability to visualize numeric, factor, and date variables effectively.
```

**DVS-2: I use plot modifications to make my visualization clear to the reader.**

-   I can modify my plot theme to be more readable

```{r}
#| label: dvs-2-ex-1

# Challenge 2 - Spicy Option

library(viridis)

ggplot(data = surveys,
        mapping = aes (x = weight, y = species, fill = sex)) +
      geom_jitter(color = "lightblue", alpha = 0.3) +
      geom_boxplot(outlier.shape = NA) +
      labs(
        title = "Side-by-Side Boxplots with Observations of Body Weight by Species",
        x = "Body Weight (grams)",
        y = "Species"
) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(family = "Times New Roman"),
        plot.title = element_text(size = 16, face = "bold"),
        panel.grid.minor = element_blank()
        )

# What I did:
# I adjusted axis text angles and font families to improve plot readability. 
# These theme customizations addressed specific feedback about text orientation and styling. 
```

-   I can modify my colors to be accessible to anyone's eyes

```{r}
#| label: dvs-2-ex-2

# Midterm comment: color-blind friendly? 

# Challenge 2 

library(viridis)

ggplot(data = surveys,
        mapping = aes (x = weight, y = species, fill = sex)) +
      geom_jitter(color = "lightblue", alpha = 0.3) +
      geom_boxplot(outlier.shape = NA) +
      scale_fill_viridis_d()

# MODIFIED

ggplot(surveys, aes(x = weight, y = species, fill = sex)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.6, width = 0.2) +
  scale_fill_viridis_d() +
  labs(title = "Weight by Species and Sex",
       x = "Weight",
       y = "Species",
       fill = "Sex") +
  theme_minimal()

# What I did:
# Made this plot color-blind friendly using the viridis palette. 
# Light blue points overlaid on filled boxplots may not have enough contrast for all viewers, especially if the boxplots have similar colors.
```

-   I can modify my plot titles to clearly communicate the data context

```{r}
#| label: dvs-2-ex-3

# Lab 2 Question 8

ggplot(data = surveys,
        mapping = aes (x = species, y = weight, fill = sex)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(color = "steelblue", alpha = 0.1) +
      labs(
        title = "Side-by-Side Boxplots with Observations of Body Weight by Species",
        x = "Species",
        y = "Body Weight (grams)"
      )
```

-   I can modify the text in my plot to be more readable

```{r}
#| label: dvs-2-ex-4

# Lab 2 Question 8

ggplot(data = surveys,
        mapping = aes (x = species, y = weight)) +
      geom_boxplot() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(family = "Times New Roman"),
        plot.title = element_text(size = 16, face = "bold")
        )
```

-   I can reorder my legend to align with the colors in my plot

```{r}
#| label: dvs-2-ex-5

# Lab 4 Question 7 

# Growing comment: Nice work reordering your legend and your facets!!!
# Do you need to rotate the x-axis labels? Is there a way you can make this plot without making people tilt their head to read the titles of your plot?

ggplot(data = ca_childcare_long,
       aes(x = study_year, y = median_price, 
           color = fct_reorder2(region, study_year, median_price))) +
  geom_point() +
  facet_grid(cols = vars(age_group)) +
  geom_smooth(method = "loess") +
  scale_x_continuous(breaks = seq(2008, 2018, 2)) +
  scale_y_continuous(breaks = seq(0, 500, by = 100)) +
  labs(x = "Study Year",
       y = " ",
       subtitle = "Weekly Median Price for Center-Based Childcare ($)",
       color = "California Region") +
  theme_bw() +
  theme(aspect.ratio = 1,                     
        axis.text.x = element_text(angle = 45, hjust = 1))

# REVISION - Remove angle to prevent head tilting

ggplot(data = ca_childcare_long,
       aes(x = study_year, y = median_price, 
           color = fct_reorder2(region, study_year, median_price))) +
  geom_point() +
  facet_grid(cols = vars(age_group)) +
  geom_smooth(method = "loess") +
  scale_x_continuous(breaks = seq(2008, 2018, 2)) +
  scale_y_continuous(breaks = seq(0, 500, by = 100)) +
  labs(x = "Study Year",
       y = " ",
       subtitle = "Weekly Median Price for Center-Based Childcare ($)",
       color = "California Region") +
  theme_bw() +
  theme(aspect.ratio = 1)
```

**DVS-3: I show creativity in my visualizations**

-   I can use non-standard colors (Example 1)

```{r}
#| label: dvs-3-1-ex-1

# Challenge 3 Question 2

ggplot(data = teacher_evals_compare,
       mapping = aes(x = sen_level, fill = set_level)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("standard" = "#D2B48C", "excellent" = '#B19CD9'))


# What I did:
# I used custom hex color codes (#D2B48C and #B19CD9) instead of default ggplot2 colors.
# These specific color choices created a distinctive visual identity for the plot. 
# This demonstrates intentional color palette selection.
```

-   I can use non-standard colors (Example 2)

```{r}
#| label: dvs-3-1-ex-2

# Lab 8 Question 4

library(RColorBrewer)

# asked ChatGPT the specifics of the brewer package
# https://chatgpt.com/share/691223ff-2670-800c-8b4c-eca3cc1c40d7

style_plot <- function(plot, 
                       theme_fn = theme_minimal, 
                       palette = brewer.pal(8, "Set2"),
                       aesthetic = "fill") {
  
  styled <- plot + theme_fn()
  
  if (aesthetic == "fill") {
    styled <- styled + scale_fill_manual(values = palette)
  } else if (aesthetic == "color") {
    styled <- styled + scale_color_manual(values = palette)
  }
  
  return(styled)
}

# What I did
# I created a reusable function that applies consistent, non-standard color schemes to any ggplot. 
# Using RColorBrewer's "Set2" palette provided eight distinct, colorblind-friendly colors. # This shows systematic approach to color theming across multiple visualizations.
```

-   I can use annotations (e.g., `geom_text()`)

```{r}
#| label: dvs-3-2

# Challenge 2 

# used ggtext's element_markdown() for colored subtitle text
ggplot(data = surveys,
       mapping = aes(x = weight, y = species, fill = sex)) +
  geom_jitter(color = "lightblue", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA) +
  labs(
    title = "Side-by-Side Boxplots with Observations of Body Weight by Species",
    subtitle = "Separated by Sex: <span style='color:#F8766D;'>Female</span> | <span style='color:#00BFC4;'>Male</span>",
    x = "Body Weight (grams)",
    y = "Species"
  ) +
  theme(plot.subtitle = element_markdown(size = 12))


# MODIFIED to use geom_text() 

# geom_text from https://r4ds.hadley.nz/communication.html#annotations

ggplot(data = surveys,
       mapping = aes(x = weight, y = species, fill = sex)) +
  geom_jitter(color = "lightblue", alpha = 0.3) +
  geom_boxplot(outlier.shape = NA) +
  geom_text(data = survey_medians,
            aes(x = median_wt, y = species, label = label),
            inherit.aes = FALSE,
            hjust = -0.2, 
            size = 3,
            color = "darkblue",
            fontface = "bold") +
  labs(
    title = "Side-by-Side Boxplots with Observations of Body Weight by Species",
    subtitle = "Separated by Sex: <span style='color:#F8766D;'>Female</span> | <span style='color:#00BFC4;'>Male</span>",
    x = "Body Weight (grams)",
    y = "Species"
  ) +
  theme(plot.subtitle = element_markdown(size = 12))



# The geom_text() modification works by adding annotations directly onto the plot to highlight specific values, in this case the median weights for each species. It uses a separate dataset (survey_medians) and maps x, y, and label independently with inherit.aes = FALSE, so it doesn’t interfere with the main plot’s aesthetics. The other parameters control the placement and appearance of the text. This satisfies the requirement by showing creative use of annotations to make the plot more informative.
```

-   I can choose creative geometries (e.g., `geom_segment()`, `geom_ribbon)()`)

```{r}
#| label: dvs-3-3

# Lab 4 Question 7

ggplot(data = ca_childcare_long,
       aes(x = study_year, y = median_price, 
           color = fct_reorder2(region, study_year, median_price))) +
  geom_point() +
  facet_grid(cols = vars(age_group)) +
  geom_smooth(method = "loess")


# Midterm comment: example not non-standard
# REVISION to use geom_segment():

# geom_segment from https://r4ds.hadley.nz/communication.html#annotations

ca_segments <- ca_childcare_long |>
  filter(study_year %in% c(2008, 2018)) |>
  group_by(region, age_group) |>
  summarise(
    start_price = median_price[study_year == 2008],
    end_price = median_price[study_year == 2018]
  )

ggplot(
  data = ca_childcare_long,
  aes(
    x = study_year,
    y = median_price,
    color = fct_reorder2(region, study_year, median_price)
  )
) +
  geom_point() +
  facet_grid(cols = vars(age_group)) +
  geom_segment(
    data = ca_segments,
    aes(
      x = 2008,
      xend = 2018,
      y = start_price,
      yend = end_price,
      color = fct_reorder2(region, 2018, end_price)
    ),
    linewidth = 1
  )
)

  
# What I did: 
# ca_segments is derived from my ca_childcare_long dataset by filtering to 2008 and 2018 and reshaping the data so each row contains both a start and end median childcare price.
# I enhanced a time series plot with geom_segment() showing changes between specific years
# These directional segments highlighted temporal trends more clearly than points alone. 
```

**DVS-4: I can calculate numerical summaries of variables.**

-   Example using `summarize()`

```{r}
#| label: dvs-4-summarize

# Lab 4 Question 5 - Using summarize for income statistics

ca_childcare |>
  filter(study_year %in% c(2008, 2018)) |>
  group_by(region, study_year) |>
  summarise(median_income = median(mhi_2018, na.rm = TRUE)) |>
  pivot_wider(names_from = study_year, values_from = median_income) |>
  arrange(desc(`2018`))
```

-   Example using `across()` - MUST BE UNIQUE EXAMPLE

```{r}
#| label: dvs-4-across

# Lab 7 Question 3 - summarizing the number of missing fishes

missing <- fish %>%
  summarise(
    rows_with_any_NA = sum(if_any(everything(), is.na)),
    across(everything(), ~sum(is.na(.)), .names = "missing_{.col}")
  )
```

**DVS-5: I can find summaries of variables across multiple groups.**

-   Example 1

```{r}
#| label: dvs-5-1

# Lab 4 Question 6

# GROWING Comment: Do you need %in% for one value? Do you need to group_by study year when there is only one year? Can you collapse your two step process (arrange + slice) into one step?

ca_childcare |>
  filter(study_year %in% c(2018)) |>
  mutate(mc_infant = as.numeric(mc_infant)) |> 
  group_by(region, study_year) |>
  summarise(median_price = median(mc_infant, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = study_year, values_from = median_price, names_prefix = "year_") |>
  arrange("year_2018") |> 
  slice_min(year_2018)

# REVISED

ca_childcare |>
  filter(study_year == 2018) |>
  mutate(mc_infant = as.numeric(mc_infant)) |> 
  group_by(region) |>
  summarise(median_price = median(mc_infant, na.rm = TRUE)) |>
  slice_min(median_price, n = 1)


# I fixed three issues in this code based on the feedback. First, I changed %in% c(2018) to == 2018 for single value filtering. Second, I removed study_year from the group_by() since we're only analyzing one year. Third, I replaced the arrange() |> slice() pattern with a single slice_min() call for cleaner code.
```

-   Example 2

```{r}
#| label: dvs-5-2

# Challenge 4

summary_table <- bay_area_long |>
  group_by(age_group, childcare_type) |>
  summarise(
    mean_cost = mean(median_price, na.rm = TRUE),
    median_cost = median(median_price, na.rm = TRUE),
    min_cost = min(median_price, na.rm = TRUE),
    max_cost = max(median_price, na.rm = TRUE),
    median_pct_families_of_color = median(pct_families_of_color, na.rm = TRUE),
    median_pct_white_families = median(pct_white_families, na.rm = TRUE),
    .groups = "drop"
  )

# What I did:
# I created a comprehensive summary table of childcare costs across age groups and provider types. 
# Multiple summary statistics provided a complete picture of cost distributions
# This shows multi-dimensional group summarization.
```

**DVS-6: I can create tables which make my summaries clear to the reader.**

-   I can modify my column names to clearly communicate the data context

```{r}
#| label: dvs-6-ex-1

# Lab 9 Question 3

# read this to help with kable styling https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html#Column__Row_Specification

surveys |>  
  map_chr(typeof) |>  
  enframe(name = "Variable", value = "Data Type") |> 
  arrange(`Data Type`, Variable) |>  
  kable(caption = "Data Types of Variables in Surveys Dataset",
        col.names = c("Variable", "Data Type")) |>  
  kable_styling(bootstrap_options = "striped") |> 
  row_spec(0, bold = TRUE)

# What I did:
# I used col.names in kable() to provide clearer, more descriptive column headers.
# Renaming technical variable names to reader-friendly terms improved table interpretability.
```

-   I can modify the text in my table to be more readable (e.g., bold face for column headers)

```{r}
#| label: dvs-6-ex-2

# Lab 9 Question 3 

# read this to help with kable styling https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html#Column__Row_Specification

surveys |>  
  map_chr(typeof) |>  
  enframe(name = "Variable", value = "Data Type") |> 
  arrange(`Data Type`, Variable) |>  
  kable(caption = "Data Types of Variables in Surveys Dataset",
        col.names = c("Variable", "Data Type")) |>  
  kable_styling(bootstrap_options = "striped") |> 
  row_spec(0, bold = TRUE)

# What I did: 
# I applied bold formatting to column headers using row_spec() for visual emphasis. 
# This simple formatting change significantly improved table readability and professional appearance. This shows attention to typographic hierarchy in tables.
```

-   I can arrange my table to have an intuitive ordering

```{r}
#| label: dvs-6-ex-3

# Lab 4 Question 5 - Arranging table by values

ca_childcare |>
  filter(study_year %in% c(2008, 2018)) |>
  group_by(region, study_year) |>
  summarise(median_income = median(mhi_2018, na.rm = TRUE)) |>
  pivot_wider(names_from = study_year, values_from = median_income) |>
  arrange(desc(`2018`))


# What I did:
# I arranged regions in descending order by 2018 median income to highlight patterns.
# Logical ordering helps readers quickly identify high and low performing groups. 
# This demonstrates thoughtful data presentation through sorting.
```

**DVS-7: I show creativity in my tables.**

-   I can use non-default colors

```{r}
#| label: dvs-7-ex-1

# Lab 9 Question 8 - Color coding missing values

library(gt)

# used this for tab_header() https://gt.rstudio.com/reference/tab_header.html
fish |> 
  map_int(~sum(is.na(.x))) |> 
  enframe(name = "Variable", value = "Missing Values") |> 
  gt() |>
  data_color(
    columns = `Missing Values`,
    fn = function(x) {
      ifelse(
        x == 0,
        "#2E8B57",   # green for not missing
        "#CD5C5C"    # red for missing
      )
    }
  ) |>
  tab_header(
    title = "Missing Values in Fish Dataset"
  )

# What I did:
# I applied conditional color coding to highlight missing data patterns in a dataset.
# Green indicated complete variables while red flagged variables with missing values. 
# This visual encoding made data quality issues immediately apparent.
```

-   I can modify the layout of my table to be more readable (e.g., `pivot_longer()` or `pivot_wider()`)

```{r}
#| label: dvs-7-ex-2

# Lab 9 Question 5 - Using pivot_longer for better table layout

evals_new |>  
  select(academic_degree, seniority, sex) |>  
  mutate(
    seniority = case_when(
      seniority <= 4                  ~ "Junior",
      seniority >= 5 & seniority <= 8 ~ "Senior",
      seniority > 8                   ~ "Very Senior"
    ),
    academic_degree = case_when(
      academic_degree == "no_dgr" ~ "No Degree",
      academic_degree == "ma"     ~ "Masters",
      academic_degree == "dr"     ~ "Doctorate",
      academic_degree == "prof"   ~ "Tenured Professor"
    ),
    sex = case_when(
      sex == "female" ~ "Female",
      sex == "male"   ~ "Male"
    )
  ) |>
  
  pivot_longer(
    cols = everything(),
    names_to = "Demographic",
    values_to = "Group"
  ) |>  
  
  filter(!is.na(Group)) |>  
  group_by(Demographic, Group) |>  
  summarize(
    Count = n(),
    Percent = str_c(round(Count / sum(Count) * 100, 2), "%")
  ) |>  
 
  mutate(
    Demographic = factor(Demographic,
                         levels = c("academic_degree", "seniority", "sex"),
                         labels = c("Academic Degree", "Seniority", "Sex"))
  ) |>  
  
  kable(
    col.names = c("Demographic", "Group", "Count", "%"),
    caption = "Demographics of Professors at the University of Poland, Warsaw in 2020"
  ) |>  
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>  
  row_spec(0, bold = TRUE) |>  
  pack_rows("Academic Degree", 1, 4) |>  
  pack_rows("Seniority", 5, 7) |>  
  pack_rows("Sex", 8, 9) |>  
  add_header_above(c(" " = 2, "Professor" = 2))


# What I did:
# I transformed wide demographic data into a long format better suited for summary presentation. 
# Using pivot_longer() created a more readable table structure with clear grouping. 
# This demonstrates data reshaping specifically for presentation purposes.
```

## Program Efficiency

**PE-1: I can write concise code which does not repeat itself.**

-   using a single function call with multiple inputs (rather than multiple function calls)

```{r}
#| label: pe-1-one-call

# Lab 4 Question 4 

ca_childcare <- ca_childcare |> 
  mutate(
    county_name = str_remove(county_name, " County"),
    region = fct_collapse(county_name,
      "Superior California" = c("Butte", "Colusa", "El Dorado", "Glenn", "Lassen", "Modoc", "Nevada", "Placer", "Plumas", "Sacramento", "Shasta", "Sierra", "Siskiyou", "Sutter", "Tehama", "Yolo", "Yuba"),
      "North Coast" = c("Del Norte", "Humboldt", "Lake", "Mendocino", "Napa", "Sonoma", "Trinity"),
      "San Francisco Bay Area" = c("Alameda", "Contra Costa", "Marin", "San Francisco", "San Mateo", "Santa Clara", "Solano"),
      "Northern San Joaquin Valley" = c("Alpine", "Amador", "Calaveras", "Madera", "Mariposa", "Merced", "Mono", "San Joaquin", "Stanislaus", "Tuolumne"),
      "Central Coast" = c("Monterey", "San Benito", "San Luis Obispo", "Santa Barbara", "Santa Cruz", "Ventura")
    )
  )

# What I did:
# I combined multiple variable creation operations into a single mutate() call. 
# This eliminated redundant function calls and made the code more efficient. 
# The approach demonstrates consolidation of similar operations.
```

-   using `across()` - MUST BE UNIQUE EXAMPLE

```{r}
#| label: pe-1-across

# Lab 8 Question 1 

rescale_column <- function(df, cols){
  df |>
    mutate(across({{cols}}, rescale_01))
}


# What I did:
# I used across() to apply the same rescaling transformation to multiple columns simultaneously. 
# This avoided repetitive code for each individual column. 
# This shows efficient bulk operation application.
```

-   using functions from the `map()` family - MUST BE UNIQUE EXAMPLE

```{r}
#| label: pe-1-map-1

# Lab 9 Question 2 - Using map_chr for data types

# used this to learn about the enframe() function https://search.r-project.org/CRAN/refmans/tibble/html/enframe.html

surveys |>  
  map_chr(typeof) |>  
  enframe(name = "Variable", value = "Data Type") |>  
  kable()


# What I did:
# I used map_chr() to efficiently determine data types for all variables in a dataset. 
# This functional approach replaced what would require manual checking of each column. 
```

**PE-2: I can write functions to reduce repetition in my code.**

-   Example 1: Function that operates on vectors

```{r}
#| label: pe-2-1

# Lab 7 Question 4 - rescale_01 function

rescale_01 <- function(x) {
  
  # if() from Base R cheatsheet https://rstudio.github.io/cheatsheets/base-r.pdf
  
  if (!is.numeric(x)) {
    return("Input must be numeric")
  }
  if (length(x) <= 1) {
    return("Input vector length must be greater than 1")
  }
  
  
  # rng() from textbook (25.2)
  rng <- range(x, na.rm = TRUE)
  min_val <- rng[1]
  max_val <- rng[2]
  
  (x - min_val) / (max_val - min_val)
}


# What I did:
# I created a reusable rescaling function that transforms any numeric vector to a 0-1 range. 
# The function includes input validation to handle edge cases gracefully. 
```

-   Example 2: Function that operates on data frames

```{r}
#| label: pe-2-2

# Lab 8 Question 1 - rescale_column function


rescale_column <- function(df, cols){
  df |>
    mutate(across({{cols}}, rescale_01))
}

rescale_column(penguins, 
               cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)
               )


# What I did
# I wrote a function that rescales specified columns within any dataframe using the across() pattern. 
# This reusable tool applies consistent transformations across multiple datasets. 
# This shows dataframe-oriented function design.
```

-   Example 3: Function that operates on vectors *or* data frames

```{r}
#| label: pe-2-3

# Lab 8 Question 2 - pivot_table function

pivot_table <- function(df, row, col){
  df |> 
    count({{row}}, {{col}}) |>
    pivot_wider(names_from = {{col}},
                values_from = n, 
                values_fill = 0) |>
    janitor::adorn_totals(c("row", "col"))
}

# What I did: 
# I created a pivot table function that works with any dataframe and column specifications. 
# The function generates formatted contingency tables with totals automatically calculated. 
```

**PE-3:I can use iteration to reduce repetition in my code.**

-   using `across()` - MUST BE UNIQUE EXAMPLE

```{r}
#| label: pe-3-across

# Lab 9 Question 4 

# ORIGINAL 
evals_new <- evals |>  
  map_if(is.character, as.factor) |>  
  bind_cols()  

# MODIFIED - use across() instead of map_if()

# where() from https://r4ds.hadley.nz/iteration.html#sec-across

evals_new <- evals |>  
  mutate(across(
    .cols = where(is.character),  # Select only character columns
    .fns = as.factor              # Convert them to factors
  ))

# What I did:
# I replaced the original map_if() approach with mutate(across()) to convert all character columns to factors.
# Using across() with where(is.character) makes the code more readable and integrates smoothly with the dplyr pipeline. 
# This demonstrates iteration to apply a function to multiple columns at once, reducing repetitive code. 
```

-   using a `map()` function with **one** input (e.g., `map()`, `map_chr()`, `map_dbl()`, etc.) - MUST BE UNIQUE EXAMPLE

```{r}
#| label: pe-3-map-1

# Lab 10 Question 1 

randomBabies <- function(n_babies){
  original <- 1:n_babies
  shuffled <- sample(original)
  correct <- sum(original == shuffled)
  return (correct)
}

results <- map_int(.x = 1:10000,
                   .f = ~randomBabies(n_babies = 4)
                  )


# What I did:
# I used map_int() to run 10,000 simulations of the random babies problem efficiently. 
# This functional approach avoided explicit loop syntax while maintaining clarity. 
```

-   using a `map()` function with **more than one** input (e.g., `map_2()` or `pmap()`) - MUST BE UNIQUE EXAMPLE

```{r}
#| label: pe-3-map-2


# Lab 10 Question 10 - Using map with function arguments

ci_dat <- map(.x = 1:1000,
              .f = ~mycifun(beta0 = 3, beta1 = 0.5, n = 100)
              )

head(ci_dat, 3)


# What I did:
# I used map() to generate confidence intervals across multiple simulation runs with different parameters. 
# This approach systematically varied function inputs while collecting results
```

**PE-4: I can use modern tools when carrying out my analysis.**

-   I can use functions which are not superseded or deprecated

```{r}
#| label: pe-4-1

# Challenge 3 Question 2 

ggplot(data = teacher_evals_compare,
       mapping = aes(x = sen_level,
                     fill = set_level)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = label_percent())

# What I did:
# I updated code from deprecated percent_format() to current label_percent() function. 
# Keeping current with package evolution ensures code longevity and compatibility. 
# This shows maintenance of modern coding practices.
```

-   I can connect a data wrangling pipeline into a `ggplot()`

```{r}
#| label: pe-4-2

# Lab 8 - Piping data manipulation into plot

penguin_boxplot(penguins, x_var = species, y_var = body_mass_g) |> 
  style_plot(theme_fn = theme_classic,
             palette = brewer.pal(3, "Pastel1"),
             aesthetic = "fill")
```

## Data Simulation & Statisical Models

**DSSM-1: I can simulate data from a *variety* of probability models.**

-   Example 1

```{r}
#| label: dsm-1-1

# Lab 10 Question 4 - Simulating data from uniform and normal distributions

intercept = 2
slope = 1

# generate x vector
x <- runif(n = 100, min = 0, max = 1)

# generate noise `ep` vector
ep <- rnorm(n = 100, mean = 0, sd = 1)

# generate outcome from population model
y = intercept + x*slope + ep


# create an "observed data" dataframe with only the x and y vectors
observed_data <- tibble(x = x, y = y)

head(observed_data)

```

-   Example 2

```{r}
#| label: dsm-1-2

# Lab 10 Question 1 - Simulating random baby assignment

randomBabies <- function(n_babies){
  original <- 1:n_babies
  shuffled <- sample(original)
  correct <- sum(original == shuffled)
  return (correct)
}

results <- map_int(.x = 1:10000,
                   .f = ~randomBabies(n_babies = 4)
                  )
```

**DSSM-2: I can conduct common statistical analyses in R.**

-   Example 1

```{r}
#| label: dsm-2-1

# Lab 2 Question 17 - Used ANOVA

# SUCCESS Comment: Do you need to save an object here? All I want is to see the output...

species_mod <- aov(weight ~ species, data = surveys)
summary(species_mod)



# REVISED

# I simplified the code by removing the unnecessary object assignment (species_mod <-) and calling summary() directly on the ANOVA result. This makes the code more concise while still providing the same statistical output, following the feedback to focus on displaying results rather than storing intermediate objects.

summary(aov(weight ~ species, data = surveys))

```

-   Example 2

```{r}
#| label: dsm-2-2

# Challenge 3 Question 3 - Used Chi-square test

# SUCCESS Comment: This works, but did you need to make a contingency table? Could you have input these vectors directly into the chisq.test() function?

chi_square <- table(teacher_evals_compare$set_level,
              teacher_evals_compare$sen_level)
chisq.test(chi_square)

# REVISED

# I simplified the code by removing the unnecessary intermediate contingency table and passing the vectors directly to chisq.test(). This makes the code more concise and efficient while producing the same statistical results, following the feedback to use the function's built-in capability to handle vector inputs.

chisq.test(teacher_evals_compare$set_level, teacher_evals_compare$sen_level)
```

-   Example 3

```{r}
#| label: dsm-2-3

# Lab 4 Question 9 - Linear regression

reg_mod1 <- lm(mc_infant ~ mhi_2018, ca_childcare)
summary(reg_mod1)
```

## Revising My Thinking

<!-- How did you revise your thinking throughout the course? How did you revise your thinking on the code examples you have provided in your portfolio? -->

<!-- For the revisions included in your Portfolio, to help me understand the nature of your revisions, please denote somehow the feedback I provided you (e.g., boldface, italics, colored text) before your revisions. -->

**Where have you revised your code?**

I revised code throughout my portfolio based on specific feedback, including simplifying filter logic, combining multiple mutate() calls and using more specific functions like slice_min() instead of arrange \|\> slice() patterns. I especially submitted revisions on Canvas throughout the quarter, going beyond the visible portfolio examples to include additional corrections on labs, challenges, and the midterm based on detailed feedback received after each submission.

**What have you learned by doing revisions?**

Revisions taught me that good code balances correctness with clarity and adaptability, and that implementing feedback develops professional coding practices while proper citation maintains academic integrity.

**How have you incorporated this knowledge into future assignments?**

After learning from revisions, I applied these lessons proactively across multiple labs and challenges. In **Lab 8** on data frame functions, I designed `style_plot()` with built-in error handling and clear parameter documentation, anticipating potential issues before they occurred. For **Lab 9** on exploratory data analysis, I consistently used `across()` with proper column selection to all summary operations. In **Challenge 4**, I implemented the feedback about combining `mutate()` calls immediately, creating cleaner pipelines for the childcare cost analysis. I also began citing resources like the R4DS textbook and official package documentation directly in my code comments, making my methods transparent and reproducible. This systematic application of revision lessons transformed how I approached coding tasks, moving from reactive corrections to proactive best practices implementation.

## Extending My Thinking

<!-- How did you extend your thinking throughout the course? How did you extend your thinking on the code examples you have provided in your portfolio? -->

**On what assignments did you push yourself by extending beyond what we learned in class?**

I particularly pushed myself in Challenge 4 with the childcare costs analysis. While we had learned the technical skills of data manipulation and visualization in class, I extended my thinking by formulating my own research question about demographic patterns in childcare access. I connected technical data analysis with real-world social equity concerns, examining how the proportion of families of color related to childcare costs across different care types.

**In what way?**

In Challenge 4, I extended my thinking by integrating multiple technical skills into a cohesive research project. I used data pivoting from Lab 4, visualization techniques from Lab 2, and demographic analysis to create a comprehensive study that connected data patterns to broader social contexts. I calculated custom metrics like proportions of families of color and conducted comparative analyses across multiple dimensions simultaneously.

Throughout the course, I consistently aimed for professional-quality outputs, paying attention to design elements, clear communication, and real-world applicability rather than just technical correctness.

## Peer Support & Collaboration

<!-- Include an image or a description of feedback you gave that you are proud of (either in a peer review or in Discord). -->

"Hello! Your code for Lab 9 is excellent and demonstrates strong mastery of the map() family functions and table formatting. Your use of map_chr() and map_int() with enframe() is clean and produces well-structured data frames that are easy to pipe into table functions.\

You did an excellent job reproducing all the target tables. Your outputs match the specifications precisely, which shows careful attention to the instructions. The gt tables are particularly polished and professional.\

I especially like your conditional coloring implementation in Question 8 using data_color() with a custom function. The green/red scheme makes missing values immediately apparent. Your multi-level grouping table in Question 6 is impressive, with row groups, spanning headers, and clean percentage formatting all working together nicely. Consider adding comments to explain complex sections like the row grouping logic in Question 6, which would help others (and you) understand the approach more quickly.\

Otherwise, your tables are good with thoughtful captions, subtitles, and formatting throughout. Great work on this assignment!"

<!-- Include a description of how you grew as a collaborator through the weekly pair programming activities.   -->

The weekly pair programming activities transformed how I approach teamwork in data science. I really enjoyed the jewel heist activity since it perfectly demonstrated how collaboration creates better solutions than working alone.

Through these sessions, I connected with more classmates and learned to work effectively under time pressure. I developed skills in clear communication, active listening, and adapting to different working styles. The experience taught me to balance confidence in my own ideas with openness to others' approaches.

These activities shifted my perspective from seeing data science as individual work to understanding it as a collaborative field where diverse perspectives create stronger, more innovative outcomes.
